---
title: Mapping literacy in Nigeria
author: Chris Littleboy
date: '2020-11-25'
slug: mapping-literacy-in-nigeria
categories:
  - Research
tags:
  - Education
  - Maps
  - Nigeria
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(knitr::opts_chunk$set(echo = F, eval = T, warning = F, message = F, comment = NA))
```

I'm beginning this blog as a way of productively procrastinating and
avoiding the task of formatting my PhD thesis. The
hope is that it turns into a space for me post short data-driven
research into various development issues. This post, for example, has
nothing to do with my research which is in agriculture, land rights and
agrarian change. But, I found the topic interesting enough to spend a
few hours making a map, and couldn't find any similar analysis
elsewhere. So I hope this is interesting and / or useful to any
potential readers!

The story behind this post is that I was having an argument with a much
better-informed friend regarding the provision of education in Nigeria.
My argument was people even in the most rural areas of Nigeria had a
much higher level of education than is often implied. Her argument, crudely put, was that public provision of primary education in Nigeria has a
lot of room for improvement. I'm an annoying person to argue with
because I don't accept an argument without seeing supporting data, and
quite reasonably nobody brings data to social engagements.

When I got home, I looked briefly for information on the regional
variation of literacy rates in Nigeria. I couldn't find anything, so
downloaded [the data](https://microdata.worldbank.org/index.php/catalog/3002/study-description) for the latest Multiple Indicator Cluster Survey
(MICS) for Nigeria which was collected in 2016/17. This contains the data of rural and urban areas of each of the 36 states in Nigeria. The questionnaire
asks members of the household if they have attended secondary school. If
they have, they are assumed literate, and if they have not they are
asked to read a short sentence. All those who were able to read the
sentence join those who attended secondary school in the group of
'literate'. Since we have information on the age and gender of each
respondent, we can get a sense of literacy rates by age, gender and
state.

So, I downloaded the data, selected the variables (literacy, age,
gender), weighted the responses, aggregated literacy rates at the state level, downloaded a [map for Nigeria](<https://gadm.org/>), and created this interactive map largely following a procedure found [here](<https://rpubs.com/castower/choroplethmap>).

Broadly, it shows that I was wrong. Literacy rates are much lower in the
North than in the South, although they are encouragingly much higher for
the younger population which shows that things are getting better. I have
been to Ondo State in South West Nigeria, where literacy is very high.
So the mistake that I made was in assuming that my personal experiences
are generalisable despite the presence of more expert opinions!

Happy to answer any questions about the data and code in the comments,
though I'd better leave the interpretation to those who know what
they're talking about...    

```{r literacymap, fig.cap = "Percentage of population that are literate. Click on states to view gender and age disaggregation. Source: MICS 2016/17."}
library(tidyverse)
library(haven)
library(raster)
library(sf)
library(ggplot2)
library(leaflet)
library(sjlabelled)
library(htmltools)
library(htmlwidgets)
select <- dplyr::select


hh <- read_sav("C:\\Users\\chris\\Downloads\\Nigeria MICS5 Datasets\\Nigeria MICS 2016-17 SPSS Datasets\\hl.sav") %>% 
  select(HH1, HH2, HH7, hhweight) %>% 
  mutate(hhid = paste0(HH1, HH2)) %>%
  select(-HH1, - HH2) %>%
  rename(state = HH7) %>%
  distinct()

men <- read_sav("C:\\Users\\chris\\Downloads\\Nigeria MICS5 Datasets\\Nigeria MICS 2016-17 SPSS Datasets\\mn.sav") %>% 
  select(HH1, HH2, MWB2,MWB7) %>%
  rename(cluster = HH1, hhid = HH2, age = MWB2, literacy = MWB7) %>%
  mutate(gender = "Male")

women <- read_sav("C:\\Users\\chris\\Downloads\\Nigeria MICS5 Datasets\\Nigeria MICS 2016-17 SPSS Datasets\\wm.sav") %>%
  select(HH1, HH2,WB2, WB7,) %>%
  rename(cluster = HH1, hhid = HH2, age = WB2, literacy = WB7) %>%
  mutate(gender = "Female")

men1524 <- men %>% 
  mutate(hhid = paste0(cluster, hhid),
         literacy = ifelse(is.na(literacy),3,literacy),
         literacy1 = ifelse(literacy == 1,1,0),
         literacy2 = ifelse(literacy == 2,1,0),
         literacy3 = ifelse(literacy == 3,1,0),
         literacy4 = ifelse(literacy == 4,1,0),
         literacy5 = ifelse(literacy == 5,1,0)) %>%
  select(-cluster) %>%
  left_join(hh) %>%
  mutate(age = ifelse(age %in% 15:24, "1524",
               ifelse(age %in% 25:35, "2535",
               ifelse(age %in% 36:49, "3649","")))) %>%
  filter(age == "1524") %>%
  mutate(w3 = literacy3 * hhweight) %>%
  group_by(state) %>%
  summarise(sum3 = sum(w3),
            sumweights = sum(hhweight)
         ) %>%
  ungroup() %>%
  mutate(literate = sum3 / sumweights) %>% 
  select(state, literate) %>%
  mutate(age = "1524", gender = "Male")

men2535 <- men %>% 
  mutate(hhid = paste0(cluster, hhid),
         literacy = ifelse(is.na(literacy),3,literacy),
         literacy1 = ifelse(literacy == 1,1,0),
         literacy2 = ifelse(literacy == 2,1,0),
         literacy3 = ifelse(literacy == 3,1,0),
         literacy4 = ifelse(literacy == 4,1,0),
         literacy5 = ifelse(literacy == 5,1,0)) %>%
  select(-cluster) %>%
  left_join(hh) %>%
  mutate(age = ifelse(age %in% 15:24, "1524",
                      ifelse(age %in% 25:35, "2535",
                             ifelse(age %in% 36:49, "3649","")))) %>%
  filter(age == "2535") %>%
  mutate(w3 = literacy3 * hhweight) %>%
  group_by(state) %>%
  summarise(sum3 = sum(w3),
    sumweights = sum(hhweight)
  ) %>%
  ungroup() %>%
  mutate(literate = sum3 / sumweights) %>% 
  select(state, literate) %>%
  mutate(age = "2535", gender = "Male")

men3649 <- men %>% 
  mutate(hhid = paste0(cluster, hhid),
         literacy = ifelse(is.na(literacy),3,literacy),
         literacy1 = ifelse(literacy == 1,1,0),
         literacy2 = ifelse(literacy == 2,1,0),
         literacy3 = ifelse(literacy == 3,1,0),
         literacy4 = ifelse(literacy == 4,1,0),
         literacy5 = ifelse(literacy == 5,1,0)) %>%
  select(-cluster) %>%
  left_join(hh) %>%
  mutate(age = ifelse(age %in% 15:24, "1524",
                      ifelse(age %in% 25:35, "2535",
                             ifelse(age %in% 36:49, "3649","")))) %>%
  filter(age == "3649") %>%
  mutate(w3 = literacy3 * hhweight) %>%
  group_by(state) %>%
  summarise(
    sum3 = sum(w3),
    sumweights = sum(hhweight)
  ) %>%
  ungroup() %>%
  mutate(literate = sum3 / sumweights) %>% 
  select(state, literate) %>%
  mutate(age = "3649", gender = "Male")

mentotal <- men %>% 
  mutate(hhid = paste0(cluster, hhid),
         literacy = ifelse(is.na(literacy),3,literacy),
         literacy1 = ifelse(literacy == 1,1,0),
         literacy2 = ifelse(literacy == 2,1,0),
         literacy3 = ifelse(literacy == 3,1,0),
         literacy4 = ifelse(literacy == 4,1,0),
         literacy5 = ifelse(literacy == 5,1,0)) %>%
  select(-cluster) %>%
  left_join(hh) %>%
  mutate(age = ifelse(age %in% 15:24, "1524",
                      ifelse(age %in% 25:35, "2535",
                             ifelse(age %in% 36:49, "3649","")))) %>%
  filter(age != "") %>%
  mutate(w3 = literacy3 * hhweight) %>%
  group_by(state) %>%
  summarise(sum3 = sum(w3),
    sumweights = sum(hhweight)
  ) %>%
  ungroup() %>%
  mutate(literate = sum3 / sumweights) %>% 
  select(state, literate) %>%
  mutate(age = "All", gender = "Male")

women1524 <- women %>% 
  mutate(hhid = paste0(cluster, hhid),
         literacy = ifelse(is.na(literacy),3,literacy),
         literacy1 = ifelse(literacy == 1,1,0),
         literacy2 = ifelse(literacy == 2,1,0),
         literacy3 = ifelse(literacy == 3,1,0),
         literacy4 = ifelse(literacy == 4,1,0),
         literacy5 = ifelse(literacy == 5,1,0)) %>%
  select(-cluster) %>%
  left_join(hh) %>%
  mutate(age = ifelse(age %in% 15:24, "1524",
                      ifelse(age %in% 25:35, "2535",
                             ifelse(age %in% 36:49, "3649","")))) %>%
  filter(age == "1524") %>%
  mutate(w3 = literacy3 * hhweight) %>%
  group_by(state) %>%
  summarise(sum3 = sum(w3),
            sumweights = sum(hhweight)
  ) %>%
  ungroup() %>%
  mutate(literate = sum3 / sumweights) %>% 
  select(state, literate) %>%
  mutate(age = "1524", gender = "Female")

women2535 <- women %>% 
  mutate(hhid = paste0(cluster, hhid),
         literacy = ifelse(is.na(literacy),3,literacy),
         literacy1 = ifelse(literacy == 1,1,0),
         literacy2 = ifelse(literacy == 2,1,0),
         literacy3 = ifelse(literacy == 3,1,0),
         literacy4 = ifelse(literacy == 4,1,0),
         literacy5 = ifelse(literacy == 5,1,0)) %>%
  select(-cluster) %>%
  left_join(hh) %>%
  mutate(age = ifelse(age %in% 15:24, "1524",
                      ifelse(age %in% 25:35, "2535",
                             ifelse(age %in% 36:49, "3649","")))) %>%
  filter(age == "2535") %>%
  mutate(w3 = literacy3 * hhweight) %>%
  group_by(state) %>%
  summarise(sum3 = sum(w3),
            sumweights = sum(hhweight)
  ) %>%
  ungroup() %>%
  mutate(literate = sum3 / sumweights) %>% 
  select(state, literate) %>%
  mutate(age = "2535", gender = "Female")

women3649 <- women %>% 
  mutate(hhid = paste0(cluster, hhid),
         literacy = ifelse(is.na(literacy),3,literacy),
         literacy1 = ifelse(literacy == 1,1,0),
         literacy2 = ifelse(literacy == 2,1,0),
         literacy3 = ifelse(literacy == 3,1,0),
         literacy4 = ifelse(literacy == 4,1,0),
         literacy5 = ifelse(literacy == 5,1,0)) %>%
  select(-cluster) %>%
  left_join(hh) %>%
  mutate(age = ifelse(age %in% 15:24, "1524",
                      ifelse(age %in% 25:35, "2535",
                             ifelse(age %in% 36:49, "3649","")))) %>%
  filter(age == "3649") %>%
  mutate(w3 = literacy3 * hhweight) %>%
  group_by(state) %>%
  summarise(
    sum3 = sum(w3),
    sumweights = sum(hhweight)
  ) %>%
  ungroup() %>%
  mutate(literate = sum3 / sumweights) %>% 
  select(state, literate) %>%
  mutate(age = "3649", gender = "Female")

womentotal <- women %>% 
  mutate(hhid = paste0(cluster, hhid),
         literacy = ifelse(is.na(literacy),3,literacy),
         literacy1 = ifelse(literacy == 1,1,0),
         literacy2 = ifelse(literacy == 2,1,0),
         literacy3 = ifelse(literacy == 3,1,0),
         literacy4 = ifelse(literacy == 4,1,0),
         literacy5 = ifelse(literacy == 5,1,0)) %>%
  select(-cluster) %>%
  left_join(hh) %>%
  mutate(age = ifelse(age %in% 15:24, "1524",
                      ifelse(age %in% 25:35, "2535",
                             ifelse(age %in% 36:49, "3649","")))) %>%
  filter(age != "") %>%
  mutate(w3 = literacy3 * hhweight) %>%
  group_by(state) %>%
  summarise(sum3 = sum(w3),
            sumweights = sum(hhweight)
  ) %>%
  ungroup() %>%
  mutate(literate = sum3 / sumweights) %>% 
  select(state, literate) %>%
  mutate(age = "All", gender = "Female")

total <- rbind(men,women) %>% 
  mutate(hhid = paste0(cluster, hhid),
         literacy = ifelse(is.na(literacy),3,literacy),
         literacy1 = ifelse(literacy == 1,1,0),
         literacy2 = ifelse(literacy == 2,1,0),
         literacy3 = ifelse(literacy == 3,1,0),
         literacy4 = ifelse(literacy == 4,1,0),
         literacy5 = ifelse(literacy == 5,1,0)) %>%
  select(-cluster) %>%
  left_join(hh) %>%
  mutate(age = ifelse(age %in% 15:24, "1524",
                      ifelse(age %in% 25:35, "2535",
                             ifelse(age %in% 36:49, "3649","")))) %>%
  filter(age != "") %>%
  mutate(w3 = literacy3 * hhweight) %>%
  group_by(state) %>%
  summarise(sum3 = sum(w3),
    sumweights = sum(hhweight)) %>%
  ungroup() %>%
  mutate(literate = sum3 / sumweights) %>% 
  select(state, literate) %>%
  mutate(age = "All", gender = "All")

literacy <- rbind(men1524,men2535,men3649,mentotal,
                  women1524,women2535,women3649,womentotal,
                  total)
state_labels = as_factor(literacy$state) %>% as_character()
literacy <- literacy %>% drop_labels() %>%
  mutate(state = state_labels) %>%
  mutate(state = ifelse(state == "FCT Abuja", "FCT", 
                 ifelse(state == "Nasarawa", "Nassarawa", state)),
         literate = round(100 * literate, digits = 0)) 

nigeria <- getData(name = "GADM", country = "NGA", level = 1)
nigeria <- st_as_sf(nigeria) %>% select(NAME_1) %>% rename(state = NAME_1) %>% 
  mutate(state = ifelse(state == "Federal Capital Territory", "FCT", state))

literacy <- literacy %>% mutate(genderage = paste0(gender, age, "age")) %>% 
  select(-age, -gender) %>%
  pivot_wider(id_cols = state, values_from = literate, names_from = genderage)
popup <- paste0("<strong>",
                "State: ",
                literacy$state, 
                "\n<strong><br /> Literacy rate: ", 
                literacy$AllAllage,
                "%",
                "\n</strong><br / > Female All: ",
                literacy$FemaleAllage,
                "%",
                "\n</strong><br />Female 15:24: ",
                literacy$Female1524age,
                "%",
                "\n</strong><br />Female 25:35: ",
                literacy$Female2535age,
                "%",
                "\n</strong><br />Female 36:49: ",
                literacy$Female3649age,
                "%",
                "\n<strong> <br />Male All: ",
                literacy$MaleAllage,
                "%",
                "</strong><br />Male 15:24: ", 
                literacy$Male1524age,
                "%",
                "\n</strong><br />Male 25:35: ",
                literacy$Male2535age,
                "%",
                "\n</strong><br />Male 36:49: ",
                literacy$Male3649age,
                "%")

literacy$popup <- popup

MapInfo <- tags$p(tags$style("<p {font-size:12px} />"),
                  tags$b("Percentage of population that are literate, defined as having attended secondary school or able to read a simple sentence. Click on states to view gender and age disaggregation. Source: MICS 2016/17."))
literacy <- merge(literacy, nigeria) %>% st_as_sf() %>% 
  st_transform(crs = "+proj=longlat +datum=WGS84") %>%
  mutate(long = unlist(st_centroid(geometry))[1],
         lat = unlist(st_centroid(geometry))[2])
pal <- colorNumeric(palette = "Blues", domain = literacy$AllAllage)

literacy_map <-
as.data.frame(literacy) %>% 
  leaflet() %>% 
  setView(lat = 9, lng =8, zoom = 5) %>%
  addProviderTiles(providers$Stamen.TonerLite) %>%
  addPolygons(data = literacy,
              popup = ~ popup,
              stroke = FALSE,
              smoothFactor = 0,
              fillOpacity = 0.7,
              color = ~ pal(literacy$AllAllage)) %>%
  addLegend("bottomright", 
            pal = pal, 
            values = ~ AllAllage,
            title = "Percentage",
            opacity = 1)

literacy_map
```
